---
title: "Tables of the raw data and estimated line means"
output: 
  workflowr::wflow_html:
    code_folding: hide 
---

```{r message=FALSE, warning=FALSE, results="hide"}
library(tidyverse)
library(DT)

# Create a function to build HTML searchable tables
my_data_table <- function(df, filename){
  datatable(
    df, rownames=FALSE,
    autoHideNavigation = TRUE,
    extensions = c("Scroller",  "Buttons"),
    options = list(
      autoWidth = TRUE,
      dom = 'Bfrtip',
      deferRender=TRUE,
      scrollX=TRUE, scrollY=1000,
      scrollCollapse=TRUE,
      buttons =
        list('pageLength', 
             list(extend = 'csv',
                  filename = filename)),
      pageLength = 10
    )
  )
}

# Load the raw data collected during the experiments (clean up in get_predicted_line_means.Rmd)
raw_female_fitness <- read_csv("data/input/female_fitness_CLEANED.csv") %>% 
  select(vial, everything()) %>% 
  rename_all(~ str_to_sentence(str_replace_all(.x, "[.]", " ")))
raw_male_fitness <- read_csv("data/input/male_fitness_CLEANED.csv") %>% 
  select(vial, everything(), -num.GFP.males) %>% 
  rename_all(~ str_to_sentence(str_replace_all(.x, "[.]", " ")))

# Load the predicted line means, as calculated in get_predicted_line_means.Rmd
predicted_line_means <- read_csv("data/derived/predicted_line_means.csv") %>% 
  mutate(female.fitness.early = round(female.fitness.early, 3),
         female.fitness.late = round(female.fitness.late, 3),
         male.fitness.early = round(male.fitness.early, 3),
         male.fitness.late = round(male.fitness.late, 3)) %>% 
  rename_all(~ str_to_sentence(str_replace_all(.x, "[.]", " "))) %>% 
  select(-Block) 
  
```

To facilitate data re-use, we here provide tables showing the raw data (i.e. the measurements of male and female fitness that were collected on each individual replicate vial), as well as the estimated line means that were calculated in [this script](get_predicted_line_means.html). Press the `CSV` button to download a particular table as a `.csv` file.

## Raw data

### Female fitness

- `Vial` is a unique ID for each vial, which groups together observations of the same group of 5 DGRP females whose fitness was measured.
- `Block` is the experimental block in which fitness was measured.
- `Line` gives the DGRP line number of the focal females.
- `Female fitness early` gives the number of L1 larvae produced by the DGRP females, in the early-life measurement.
- `Female fitness late` gives the number of L1 larvae produced by the DGRP females, in the late-life measurement.

```{r}
my_data_table(raw_female_fitness,
              "female_fitness_raw_data")
```

### Male fitness

- `Vial` is a unique ID for each vial, which groups together observations of the same group of 5 DGRP females whose fitness was measured.
- `Block` is the experimental block in which fitness was measured.
- `Line` gives the DGRP line number of the focal males
- `Early male rival` gives the number of L1 larvae sired by the rival _GFP_ males, in the early-life measurement.
- `Early male focal` gives the number of L1 larvae sired by the focal DGRP males, in the early-life measurement.
- `Late male rival` gives the number of L1 larvae sired by the rival _GFP_ males, in the late-life measurement.
- `Late male focal` gives the number of L1 larvae sired by the focal DGRP males, in the late-life measurement.

```{r}
my_data_table(raw_male_fitness,
              "male_fitness_raw_data")
```

## Estimated line means

The four columns other than `Line` give the line mean fitness for each of the four fitness traits. Fitness was estimated from Bayesian multivariate Poisson or Binomial models described in [this script](get_predicted_line_means.html). The fitness values were expressed on the scale of the linear predictor (and thus are approximately normally distributed), and have been scaled to have a mean of zero and variance of one. Thus, a line with mean fitness equal to 1.5 is 1.5 standard deviations above average, on the scale of the linear predictor. Note that male fitness was not measured in `line_354` due to experimenter error.

```{r}
my_data_table(predicted_line_means,
              "line_mean_fitness_data")
```

